---
title: "storm data"
author: "Rick Yoo"
date: '2020/1/27 '
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# <Top 10 Weather events 'most harmful to population health' & 'with greatest economic consequences'>
### Introduction
Given a set of storm data, the purpose of this research is to answer these two questions - 1) "Which types of events are most harmful to population health?" and 2) "Which types of events have the greatest economic consequences?" - in a way that is reproducible by anyone. 

# library & sessionInfo
- load necessary libraries
```{r}
suppressPackageStartupMessages({
library(R.utils)
library(data.table)
library(ggplot2)
library(dplyr)
library(cowplot)
})
sessionInfo()
```

# Data processing
 - unzip & fread
```{r cache=TRUE}
fileName <- "repdata%2Fdata%2FStormData.csv.bz2"
if (!fileName %in% list.files()) {
    fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
    download.file(fileUrl, fileName)
    bunzip2(filename = fileName, remove = FALSE)
}
# data read
DF <- fread(gsub(".bz2", "", fileName))
```

### Question 1. Across the United States, which types of events (as indicated in the \color{red}{\verb|EVTYPE|}EVTYPE variable) are most harmful with respect to population health?
 - select a few variables that are related to population health
 - group by event type
 - calculate sum of 'fatalities', 'injuries' for each event type
```{r}
PopHealth <- DF %>% 
    # select population health - related variables
    select(EVTYPE, FATALITIES, INJURIES)%>%
    # group by event type 
    group_by(EVTYPE) %>% 
    # sum of Fatalities & Injuries for each event type
    summarise(
    	totalF = sum(FATALITIES), 
    	totalI = sum(INJURIES)
    	)
```
 - sort by total fatalities & select top 10
 - create a bar plot and assign it to `plotFat`
```{r}
## by Fatalities
Fat <- PopHealth %>%
    arrange(-totalF) %>%
    head(10)
plotFat <- ggplot(data = Fat) +
	geom_bar(mapping = aes(x = reorder(EVTYPE, -totalF), y = totalF), 
	         fill = 'red', stat = 'identity') +
    labs(title = 'Fatalities - Top 10', 
         x = 'Event Type',
         y = 'Fatalities Count') +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
 - sort by total injuries & select top 10
 - create a bar plot and assign it to `plotInj`
```{r}
## by Injuries
Inj <- PopHealth %>%
    arrange(-totalI) %>%
    head(10)
plotInj <- ggplot(data = Inj) +
	geom_bar(mapping = aes(x = reorder(EVTYPE, -totalI), y = totalI), 
	         stat = 'identity', fill = 'blue') +
    labs(title = 'Injuries - Top 10', 
         x = 'Event Type',
         y = 'Injuries Count') +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
 - show two barplots in one figure using `plot_grid()` (from the cowplot library)
```{r}
# plot into one
theme_set(theme_bw())
plot_grid(plotFat, plotInj, labels='AUTO')
```

### Question 2. Across the United States, which types of events have the greatest economic consequences?
 - select a few variables that are related to economic consequences
 - multiply `PROPDMG` by appropriate scale indicated in `PROPDMGEXP`
 - group by event type
 - calculate sum of 'property damage', 'crop damage' for each event type
```{r}
Eco <- DF %>% 
    # select economic consequence - related variables
    select(EVTYPE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP) %>%
    mutate(
        PROPDMG = case_when(
            PROPDMGEXP == 'h' | PROPDMGEXP == 'H' ~ PROPDMG * 100,
            PROPDMGEXP == 'k' | PROPDMGEXP == 'K' ~ PROPDMG * 1000,
            PROPDMGEXP == 'm' | PROPDMGEXP == 'M' ~ PROPDMG * 1000000,
            PROPDMGEXP == 'b' | PROPDMGEXP == 'B' ~ PROPDMG * 1000000000,
            TRUE ~ PROPDMG
        ), 
        CROPDMG = case_when(
            CROPDMGEXP == 'h' | CROPDMGEXP == 'H' ~ CROPDMG * 100,
            CROPDMGEXP == 'k' | CROPDMGEXP == 'K' ~ CROPDMG * 1000,
            CROPDMGEXP == 'm' | CROPDMGEXP == 'M' ~ CROPDMG * 1000000,
            CROPDMGEXP == 'b' | CROPDMGEXP == 'B' ~ CROPDMG * 1000000000,
            TRUE ~ CROPDMG
        )
    ) %>%
    # group by event type 
    group_by(EVTYPE) %>% 
    # sum of PROP & CROP Damage for each event type
    summarise(
    	totalP = sum(PROPDMG), 
    	totalC = sum(CROPDMG)
    	)
```
 - sort by total Property damage & select top 10
 - create a bar plot and assign it to `plotProp`
```{r}
## by PROP
Prop <- Eco %>%
    arrange(-totalP) %>%
    head(10)
plotProp <- ggplot(data = Prop) +
	geom_bar(mapping = aes(x = reorder(EVTYPE, -totalP), y = totalP), 
	         fill = 'red', stat = 'identity') +
    labs(title = 'Property Damage - Top 10', 
         x = 'Event Type',
         y = 'Property Damage ($)') +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
 - sort by total Crop damage & select top 10
 - create a bar plot and assign it to `plotCrop`
```{r}
## by CROP
Crop <- Eco %>%
    arrange(-totalC) %>%
    head(10)
plotCrop <- ggplot(data = Crop) +
	geom_bar(mapping = aes(x = reorder(EVTYPE, -totalC), y = totalC), 
	         fill = 'blue', stat = 'identity') +
    labs(title = 'Crop Damage - Top 10', 
         x = 'Event Type',
         y = 'Crop Damage ($)') +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
 - show two barplots in one figure using `plot_grid()` (from the cowplot library)
```{r}
plot_grid(plotProp, plotCrop, labels='AUTO')
```

# Results
### 1. Top 10 events considering 'Fatalities' and 'Injuries'
```{r}
print(Fat)
print(Inj)
```
### 2. Top 10 events considering 'Property damage' and 'Crop damage'
```{r}
print(Prop)
print(Crop)
```


